---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# afpca

<!-- badges: start -->

[![R-CMD-check](https://github.com/angelgar/afpca/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/angelgar/afpca/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Adaptive Functional Principal Component Analysis

* Authors: [Angel Garcia de la Garza](http://angelgarciadelagarza.com), [Britton Sauerbrei](https://sauerbreilab.org/), [Jeff Goldsmith](https://jeffgoldsmith.com/)
* License: [MIT](https://opensource.org/licenses/MIT). See the [LICENSE](LICENSE) file for details
* Version: 0.9

## What is does

afpca implements Adaptive Functional Principal Component Analysis (aFPCA), a method for estimating directions of variation in functional data that exhibit sharp changes in smoothness. Standard FPCA methods impose a global smoothness assumption that can fail to capture abrupt transitions in the underlying signal. afpca addresses this by combining a fast and scalable adaptive scatterplot smoothing technique with a probabilistic FPCA framework, allowing functional principal components to be smoothed adaptively. This is particularly useful in applications such as neural recordings, where sharp changes in activity following a stimulus must be distinguished from smooth baseline behavior.

## Installation

You can install the development version of afpca from [GitHub](https://github.com/) with:

``` r
install.packages("devtools")
devtools::install_github("angelgar/afpca")
```

## How to use it

These are examples of running adaptive FPCA. More details of the use of the package can be found in XYZ. 

The code below uses a the function `afpca::simulate_adaptive_functional_data()` to simulate 20 curves $Y_i(t)$, $i = 1,\dots,20$ observed over 200 time points on a common grid over domain (0,1) with gaussian noise. These functions are generated from a mean function and two functional principal components with varying temporal smoothness defined as:

* $\mu(t) = t^{-3/2}$ $\sin(\pi \times t^{1/4})I(t>1/2)$ 
* $\phi_k(t) = t^{-3/2}$ $\sin(4\pi \times k \times t^{1/4})I(t>1/2), k = 1,2$


```{r}

library(afpca)

simulated_data <- simulate_adaptive_functional_data(N.subj = 20)

```

The plot below show what this simulated data looks like:

```{r, simulated_data_plot, echo = FALSE, message= FALSE, warning=FALSE, fig.align='center', fig.height=3, fig.width=7, dpi = 300}

library(tidyverse)

simulated_data$data %>%
  as_tibble() %>%
  mutate(time = 1:200) %>%
  pivot_longer(-time) %>%
  ggplot(aes(x = time, y = value)) +
    geom_line(aes(group = name), alpha = 0.4) + 
    theme_bw() + xlab("Domain") + ylab("Value")

```

Our software performs adaptively-smoothed functional principal component analysis. The main function in our package to do this is `afpca:fpca.adapt`. The code below illustrates how to do this:

```{r}

afpca.output <- fpca.adapt(data = simulated_data)

```

The plots below show the estimated mean function and functional principal components 

```{r estimated_fpc_plot, echo = FALSE, fig.align='center', fig.height=3, fig.width=9, dpi = 300}

mean_plot <- afpca.output$mean %>%
                as_tibble() %>%
                mutate(time = 1:200) %>%
                ggplot(aes(x = time, y = V1)) +
                  geom_line() + 
                  theme_bw() + 
                  labs(y = expression(paste(hat(mu),"(t)")),
                       x = "domain")


phi_plot <- afpca.output$fpcs %>%
                as_tibble() %>%
                mutate(time = 1:200) %>%
                pivot_longer(-time) %>%
                ggplot(aes(x = time, y = value)) +
                  geom_line(aes(color = name)) + 
                  theme_bw() + 
                  labs(y = expression(paste(hat(mu),"(t)")),
                       x = "domain") + 
                  scale_color_manual(values = c("red","blue"), 
                                     labels = expression(paste(hat(phi)[1],"(t)"),
                                                         paste(hat(phi)[2],"(t)")))

library(patchwork)

mean_plot + phi_plot + plot_layout(guides = 'collect')   &
  theme(legend.position = 'bottom',
        legend.direction = 'horizontal')

```


The plots below shows two examples of observed functions and the respective reconstructions. 

```{r reconstructed_plots, echo = FALSE, fig.align='center', fig.height=3, fig.width=9, dpi = 300}

example1_plot <- cbind(observed = afpca.output$Y[,1], reconstructed = afpca.output$Y_hat[,1]) %>%
                    as_tibble() %>%
                    mutate(time = 1:200) %>%
                    pivot_longer(-time) %>%
                    ggplot(aes(x = time, y = value)) +
                      geom_line(aes(color = name)) + 
                      theme_bw() + 
                      ylab("value") + xlab("domain") + 
                      scale_color_manual(values = c("red","black"), name = "Type") + ylim(-.3, 1)


example2_plot <- cbind(observed = afpca.output$Y[,7], reconstructed = afpca.output$Y_hat[,7]) %>%
                    as_tibble() %>%
                    mutate(time = 1:200) %>%
                    pivot_longer(-time) %>%
                    ggplot(aes(x = time, y = value)) +
                      geom_line(aes(color = name)) + 
                      theme_bw() + 
                      ylab("value") + xlab("domain") +  
                      scale_color_manual(values = c("red","black"), 
                                         name = "Type") + ylim(-.3, 1)


example1_plot + example2_plot + plot_layout(guides = 'collect')  &
  theme(legend.position = 'bottom',
        legend.direction = 'horizontal')


```

## Citation

Garcia de la Garza, A., Sauerbrei, B., Hantman, A., & Goldsmith, J. (2023). Adaptive Functional Principal Component Analysis. arXiv preprint arXiv:2310.01760.

