---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# afpca

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/afpca)](https://CRAN.R-project.org/package=afpca)
[![R-CMD-check](https://github.com/angelgar/afpca/workflows/R-CMD-check/badge.svg)](https://github.com/angelgar/afpca/actions)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Adaptive Functional Principal Component Analysis

* Authors: [Angel Garcia de la Garza](http://angelgarciadelagarza.com), [Britton Sauerbrei](https://sauerbreilab.org/), [Jeff Goldsmith](https://jeffgoldsmith.com/)
* License: [MIT](https://opensource.org/licenses/MIT). See the [LICENSE](LICENSE) file for details
* Version: 0.9

## What is does

TEXT TEXT

## Installation

You can install the development version of afpca from [GitHub](https://github.com/) with:

``` r
install.packages("devtools")
devtools::install_github("angelgar/afpca")
```

## How to use it

These are examples of running adaptive FPCA. More details of the use of the package can be found in XYZ. 

The code below uses a the function `afpca::simulate_adaptive_functional_data()` to simulate 20 curves $Y_i(t)$, $i = 1,\dots,20$ observed over 200 time points on a common grid over domain (0,1) with gaussian noise. These functions are generated from a mean function and two functional principal components with varying temporal smoothness defined as:

* $\mu(t) = t^{-3/2}$ $\sin(\pi \times t^{1/4})I(t>1/2)$ 
* $\phi_k(t) = t^{-3/2}$ $\sin(4\pi \times k \times t^{1/4})I(t>1/2), k = 1,2$


```{r}

library(afpca)

simulated_data <- simulate_adaptive_functional_data(noise.var = 0.05)

```

The plot below show what this simulated data looks like:

```{r, echo = FALSE, message= FALSE, warning=FALSE, fig.align='center', fig.height=4, fig.width=6}

library(tidyverse)

simulated_data$data %>%
  as_tibble() %>%
  mutate(time = 1:200) %>%
  pivot_longer(-time) %>%
  ggplot(aes(x = time, y = value)) +
    geom_line(aes(group = name), alpha = 0.4) + 
    theme_bw() + xlab("Domain") + ylab("Value")
  

```






## Citation

## References


